---
title: "Problem Set 5"
author: "Nitya Lagadapati"
date: "2023-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Load and check data (5pt)
```{r}
## Question 1
gapminder <- read.delim("gapminder.csv.bz2")

```

```{r}
## Question 2
head(gapminder,5)
```

#### 2. (2pt) Load data. How many rows/columns do we have? 

We have  `r nrow(gapminder)`  rows and `r ncol(gapminder)` columns in this data set.

#### 3. (2pt) Print a small sample of data. Does it look ok? 

The sample looks okay but it seems as though there are a lot of missing (NA) values and empty values which is something to be careful of when analyzing the data and making plots. 

## Descriptive statistics (15pt)
```{r}
## Question 1
iso3 <- n_distinct(gapminder$iso3)
name <- n_distinct(gapminder$name)
iso2 <- n_distinct(gapminder$iso2)

```
#### 1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.
 
If we filter by iso2, there are `r iso2` different countries in the data set. If we filter by name, there are `r name` different countries and by iso3, there are `r iso3` different countries in the data set. 


#### 2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more iso3 -codes. What is going on? Can you find it out?

```{r}
##Question 2a
iso2_name <- gapminder %>% select(iso2, name) %>% group_by(iso2) %>% summarize(unique_names_number = n_distinct(name, na.rm = TRUE)) %>% filter(unique_names_number > 1) 

name_iso2 <- gapminder %>% select(iso2,name) %>% filter(iso2 == "") %>% group_by(name) %>% summarize()

```

##### (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?
There is only `r nrow(iso2_name)` ISO-2 code that corresponds to more than 1 name, all the other countries have only 1 name per ISO-2 code. Although the ISO-2 code for the country corresponding to more than one name is blank, the country names it corresponds to is `r name_iso2[1,]`, represented by a blank name and `r name_iso2[2,]`. 

```{r}
## Question 2b
iso3_name <- gapminder %>% select(iso3, name) %>% group_by(name) %>% summarize(unique_names_number = n_distinct(iso3, na.rm = TRUE)) %>% filter(unique_names_number > 1) 

name_iso3 <- gapminder %>% select(iso3,name) %>% filter(name == "") %>% group_by(iso3) %>% summarize()
```

##### (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are CHANISL and NLD CURACAO.
There are `r nrow(name_iso3)`, country names corresponding to more than one ISO-3 code. These countries are, `r name_iso3[1,]`, `r name_iso3[2,]`, `r name_iso3[3,]` and, `r name_iso3[4,]`. 

```{r}
##Question 3
min_year <- min(gapminder$time, na.rm = TRUE)
max_year <- max(gapminder$time, na.rm = TRUE)
```

#### 3. (2pt) What is the minimum and maximum year in these data?
The maximum year in this data is `r max_year` and the minimum year is `r min_year`. 

## CO2 emissions (30pt)

```{r}
## Question 1
missing_co2 <- sum(is.na(gapminder$co2))

missing_co2_PC <- sum(is.na(gapminder$co2_PC))

by_year <- gapminder %>% select(time,co2,co2_PC) %>% group_by(time) %>% summarize(na_co2 = sum(is.na(co2)), na_co2_pc = sum(is.na(co2_PC))) %>% mutate(total_na = (na_co2 + na_co2_pc)) %>% select(time, total_na) %>% filter(total_na == max(total_na)) 


```

#### 1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2 and co2_PC. Which years have most missing data?
CO2 emissions has `r missing_co2` missing data points and CO2_PC has `r missing_co2_PC` missing data points. Summing up the total missing CO2 and CO2_PC values, I found that the years with the most missing data were `r by_year[1,1]`, `r by_year[2,1]`, and `r by_year[3,1]`, with total missing values of `r by_year[1,2]`. 

```{r}
## Question 2
co2_plot <- gapminder %>% select(name,time,co2) %>% filter(name == "United States of America"| name == "China"|name == "India"|name == "Russian Federation"|name == "Brazil"|name == "Sudan"|name == "France", !is.na(co2))

ggplot(co2_plot, aes(x = time, y = co2, color = name))+
  geom_line()+
   ggtitle("CO2 Emissions Over Time")+
  labs(x = "Time", y = "CO2", color = "Country")
  

```

#### 2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.
I noticed that China has the most growth in CO2 emissions over the years when compared to Brazil, France, India, Sudan, Russia, and the USA. While India and USA also have high CO2 emissions, USA seems to have a declining trend while India's emissions are growing. Brazil, France, and Sudan have low emission rates to begin with and are generally maintained over time. Russia, which once had high CO2 emission rates, declined over time starting at about 1990. These trends indicate that the world's major manufacturing countries, with high populations, such as China, India, and the USA have the largest CO2 emissions. This makes sense because when populations and industries present are high, the energy used increases, thereby increasing CO2 emissions. 

```{r}
## Question 3
co2_pc_plot <- gapminder %>% select(name,time,co2_PC) %>% filter(name == "United States of America"| name == "China"|name == "India"|name == "Russian Federation"|name == "Brazil"|name == "Sudan"|name == "France", !is.na(co2_PC))

ggplot(co2_pc_plot, aes(x = time, y = co2_PC, color = name))+
  geom_line()+
   ggtitle("CO2 Emissions Per Capita Over Time")+
  labs(x = "Time", y = "CO2 emissions Per Capita", color = "Country")

```

#### 3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the same countries. What does this figure suggest?
According to the CO2 emissions per capita plot, it is obvious that population has an impact on CO2 emissions, even though other factors may play a role as well. Because this graph accounts for the difference in population between countries, it is easier to compare the countries and their CO2 emissions over time. The graph indicates that USA and Russia have always had, and continue to have the highest CO2 emissions per capita when compared to India, Brazil, France, China, and Sudan. China and India seem to be the only countries out of all those in the plot that are beginning to increase in CO2_PC, the other countries are declining in emissions. This suggests that there is an increase in energy consumption per person in India and China, but a decrease in carbon based energy consumption in the rest of the countries. This may indicate a shift in the locations of major manufacturing industries as well as an increase in non-carbon energy sources of nations showing a decline in CO2 emissions.  

```{r}
## Question 4
co2_region_plot <- gapminder %>% select(region,time,co2_PC) %>% filter(!is.na(co2_PC),!is.na(region),!is.na(time),!region == "") %>% group_by(region,time) %>%  summarize(mean_co2 = mean(co2_PC)) 

ggplot(co2_region_plot, aes(x = time, y = mean_co2, color = region))+
  geom_line()+
   ggtitle("Average CO2 Emissions Over Time Per Capita Per Continent")+
  labs(x = "Time", y = "Average CO2 Emissions Per Capita", color = "Region")
```


#### 4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see. Note: just compute averages over countries and ignore the fact that countries are of different size. Hint: Americas 2016 should be 4.80.

According to the per region, is is obvious that Africa's CO2 emissions have not significantly changed from 1960 to 2019, and they have been the lowest out of all the regions for this entire period. The Americas on the other hand saw a significant decrease in CO2 emissions overall and saw many peaks and low points over the years. Asia and Oceania both saw in increase in CO2 but Asia saw a more significant increase implying that Asia had more CO2 emissions per person develop over this period, perhaps due to an increase in manufacturing industries using carbon based energy sources. Europe's emissions in 2016 and 1960 are relatively the same but it experiences changes over the years. Essentially, this plot indicates that there was a shift in the largest CO2 producing regions from Europe/Americas to Asia and Europe. Asia makes sense because in the present, Asia is increasing manufacturing industries and depends on carbon based energy sources especially with their oil production countries so these countries are bound to emit more CO2 per person. The Americas are switching to other sources such as light, and water so they emit less CO2, and European countries have small populations and carbon dependent energy sources so their emissions are higher per person. 

```{r}
## Question 5
co2_region_bar <- co2_region_plot %>% filter(time == "1960"| time == "2016", !region == "") 

ggplot(co2_region_bar, aes(x = region, y = mean_co2, fill = as.factor(time))) + 
  geom_col(position = "dodge2") + 
  ggtitle("Average CO2 Per Capita in 1960 vs 2016")+
  labs(x = "Regions", y = "Average CO2 per Capita", fill = "Year")
  
```

#### 5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016. 
The barplot also indicates that Asia had the largest increase in CO2 emissions from 1960 to 2016 while the Americas significantly dropped in CO2 emissions. From the line graph, I thought Africa and Europe's emissions stayed relatively the same but with the bar plot, it is obvious that Africa and Europe's emissions also increased. This is likely because the scale for this bar graph is different from the last one making it easier to see the difference between the starting and ending year's CO2 emissions. Oceania also increased in CO2 emissions. Overall, this barplot matches the line graph but makes the difference between the starting and ending year's CO2 emissions more prominent, which in turn supports the idea that there was a shift in CO2 producing industries from the Americas and Europe to Asia and Europe. 

```{r}
## Question 6
largest <- gapminder %>% select(time,name,region,co2_PC) %>% filter(!is.na(name), !name == "", !is.na(co2_PC), !is.na(region), !region == "", time == "2016") %>% group_by(region) %>% top_n(3,co2_PC) %>% arrange(region)

smallest <- gapminder %>% select(time,name,region,co2_PC) %>% filter(!is.na(name), !name == "", !is.na(co2_PC), !is.na(region), !region == "", time == "2016") %>% arrange(co2_PC) %>% group_by(region) %>%  slice(1:3) %>% group_by(region)
```

#### 6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).

Since the year 2019 had lots of missing data, the following comparisons are for 2016 instead.
In 2016, the three largest CO2 emitters for Africa are `r largest[1,2]`, `r largest[2,2]`, and `r largest[3,2]`, and the three smallest are `r smallest[1,2]`, `r smallest[2,2]`, and `r smallest [3,2]`. In 2016, the three largest CO2 emitters for the Americas are the `r largest[4,2]`, `r largest[5,2]`, and `r largest[6,2]`, and the three smallest are `r smallest[4,2]`, `r smallest[5,2]`, and `r smallest[6,2]`. In 2016, the three largest CO2 emitters for Asia are `r largest[7,2]`, `r largest[8,2]`, and `r largest[9,2]` and the three smallest are `r smallest[7,2]`, `r smallest[8,2]`, `r smallest[9,2]`. In 2016, the three largest CO2 emitters for Europe are `r largest[10,2]`, `r largest[11,2]`, and `r largest[12,2]`, and the three smallest are `r smallest[10,2]`, `r smallest[11,2]`, and `r smallest[12,2]`. In 2016, the three largest CO2 emitters for Oceania are `r largest[13,2]`, `r largest[14,2]`, and `r largest[15,2]`, and the three smallest are `r smallest[13,2]`, `r smallest[14,2]`, and `r smallest[15,2]`. 

## GDP per capita (50pt)
```{r}
## Question 1
GDP_LE_1960 <- gapminder %>% select(name,GDP_PC,lifeExpectancy,totalPopulation, time, region) %>% filter(time == "1960", !is.na(GDP_PC), !is.na(lifeExpectancy), !is.na(totalPopulation), !name =="", !is.na(region))

ggplot(GDP_LE_1960, aes(x = GDP_PC, y = lifeExpectancy, color = region))+
  geom_point(aes(size = totalPopulation))+
    ggtitle("GDP vs Life Expectancy in 1960")+
  labs(x = "GDP per Capita", y = "Life Expectancy", color = "Region", size = "Population")

```

#### 1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better.Comment what do you see there.

All of Europe and some countries in America have the largest life expectancy and highest GDP rates, which also correspond with smaller populations ( expect for some outliers). Africa, Asia, and Oceania have lower GDPs and life expectancy. This makes sense because that countries we consider "developed" in the 1960s tend to be Western countries in Europe and the Americas, and since they were the ones with more money, they were able to lead lifestyles that increased their expected lifetime. Countries in Asia, and Africa tend to have high populations and less resources per person which results in less GDP and a decrease in life expectancy because the lifestyles increasing life expectancy cannot be maintained without wealth. This same logic applies to the American nations with lower life expectancy. Although Oceania has countries with relatively small populations, they are isolated small islands without much access to resources that increase lifespan, which is why most of these countries have small life spans with the exception of the two closer to Europe. Access to medicine, healthcare, education etc, all require wealth, and they help increase life expectancy when available, so if one doesn't have money they are less likely to have a longer life because they cannot access resources.

```{r}
## Question 2
GDP_LE_2019 <- gapminder %>% select(name,GDP_PC,lifeExpectancy,totalPopulation, time, region) %>% filter(time == "2019", !is.na(GDP_PC), !is.na(lifeExpectancy), !is.na(totalPopulation), !name =="", !is.na(region))

ggplot(GDP_LE_2019, aes(x = GDP_PC, y = lifeExpectancy, color = region))+
  geom_point(aes(size = totalPopulation))+
    ggtitle("GDP vs Life Expectancy in 2019")+
  labs(x = "GDP per Capita", y = "Life Expectancy", color = "Region", size = "Population") 
  
```

#### 2. (4pt) Make a similar plot, but this time use 2019 data only.

#### 3. (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years?

In 2019, almost all countries in regions globally seemed to have increased in life expectancy and population. This is shown by the shift the of y-axis scale to include above 80 years of age, since now there are countries with life expectancy above 80 years and by the change in the legend showing larger values for population. This indicates that globally, we had an increase in resources and wealth which led to an overall growth in the lifetime length of the human population. Looking at the change, we can see that Asian countries, represented by the green dots, moved up the graph significantly more than countries in other regions. This indicates that Asian countries are developing at a faster rate, and quickly reaching that of countries in Europe and the Americas. However, Asia is an exception, countries in Africa and most in Oceania continue to lag behind Europe and the Americas and despite significant growth in Africa, the life expectancy continues to be smaller compared to other countries. Europe and the Americas didn't really change over this period in terms of life expectancy, but this makes sense since nations in these regions continued to be global superpowers 1960-2019, so they aren't going to be growing by drastic amounts in terms of life expectancy.  

```{r}
## Question 4
avg_LE <- gapminder %>% select(time,region,lifeExpectancy) %>% filter(!region == "", time == "1960"| time == "2019", !is.na(lifeExpectancy)) %>% group_by(region,time) %>% summarize(avgLE = mean(lifeExpectancy))

cat("Average Life Expectancy per region in 1960 and 2019")
head(avg_LE, 10)
```

#### 4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures? Note: here as average I mean just average over countries, ignore the fact that countries are of different size.

The average life expectancy increased for all regions according to the newly calculated table and this matches what I saw between the two graphs above. This table indicates that life expectancy increased by about 20 years on average per continent for most continents. This means most countries should have seen a similar increase in life expectancy. The scale of the y-axis in 2019, shifted about 20 years compared to 1960, so this indicates that the in general, countries on the graph also increased their lifetime by about 20 years. Therefore, the results seen in the table fit with the results in the figure. 


```{r}
## Question 5

LE_lag <- gapminder %>% select(time,region,lifeExpectancy) %>% filter(!region =="", !is.na(lifeExpectancy), time == "2019"|time =="1960") %>% group_by(region,time) %>%  summarize(mean = mean(lifeExpectancy, na.rm = TRUE)) %>% mutate(prev = lag(mean), growth = mean-prev) %>% filter(!is.na(growth)) %>% arrange(growth)
cat("Average LE Growth Per Country")
head(LE_lag,5)
```

#### 5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see. Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up! See https://faculty.washington. edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.

I assumed this question to mean the average of the growth between 1960 and 2019 per country in each region. The average life expectancy growth from 1960-2019 was the largest for Asia and the smallest for Europe with values `r LE_lag[5,5] ` years and `r LE_lag[1,5]` years respectively. The other regions fall in between these two extremes. Africa had the second highest growth with a value of `r LE_lag[4,5]`years which is almost the twice the growth of Europe. Oceania had a growth of `r LE_lag[2,5]` and the Americas a growth of `r LE_lag[3,5]` years. This indicates that Asia and Africa had a much larger life expectancy growth over this time period. This implies that countries in Asia and Africa must have had access to more resources and money over this time, than they began with, such that they could increase their lifetime. While Americas and Oceania grew, they didn't grow as much meaning they didn't have as much room to develop as Africa and Asia. And Europe, known to contain many superpower countries, grew the least because it already had a high life expectancy to begin with and didn't develop significantly like countries in Africa or Asia to show a large increase in life expectancy.  

```{r}
## Question 6
histogram <- gapminder %>% select(time,GDP_PC) %>% filter(time == "2019"|time == "1960", !is.na(GDP_PC)) 

ggplot(histogram,aes(x = GDP_PC, fill = as.factor(time))) +
  geom_histogram(position = "dodge", bins = 25)+
  ggtitle("GDP per Capita in 1960 vs 2019")+
  labs(x = "GDP per Capita", y = "Number of Countries", fill = "Year")
  
```

#### 6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!

This graph shows that there were more countries with higher GDP per Capita in 2019 than in 1960 but at the same time there were more countries with lower GDPs. This could indicate an increase in the number of countries in general over the years, or the fact that the GDPs are more spread out due to an increase in global superpowers. This would create a more prominent divide between countries of the world, meaning there are poorer countries and richer countries (essentially greater variability in data). 


```{r}
## Question 7
ranks <- gapminder %>% select(time,name,lifeExpectancy) %>% filter(!name == "", time == "1960"|time == "2019", !is.na(lifeExpectancy)) %>% group_by(time) %>% mutate(rank = rank(-lifeExpectancy)) %>% filter(name == "United States of America")
```

#### 7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.) Hint: check out the function rank()! Hint2: 17 for 1960.
The ranking of the US in terms of life expectancy is `r ranks[1,4]` in 1960 and `r ranks[2,4]` in 2019. 

```{r}
## Question 8
ranks_cont <- gapminder %>% select(time,name,lifeExpectancy) %>% filter(!name == "", time == "1960"|time == "2019", !is.na(lifeExpectancy)) %>% group_by(time) %>% mutate(rank = rank(-lifeExpectancy)/n_distinct(name)) %>%  filter(name == "United States of America")

```

#### 8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year? Hint: 0.0904 for 1960.
The relative ranking of the US in terms of life expectancy is `r ranks_cont[1,4]` in 1960 and `r ranks_cont[2,4]` in 2019. 

#### Finally tell us how many hours did you spend on this PS.
I spent 4 hours total on the problem set.
